Settler colonial projects and variation in paths to building the welfare and carceral state

### 2023-12-06 t1_analysis.Rmd
ff

# TO DO

leaving out railroad grants induces bias if morill or homestead grants are negatively correlated with railroad grants.

2. Lit review
3. writeup demographics and bivars
4. models

# Theoretical framework

HOW ARE PROJECTS / PROBLEMS OF SETTLER COLONIAL GOVERNANCE RESOLVED? WHAT ORGS AND INSTITUTIONS ARE BUILT?

Settler colonial regimes seek to resolve different problems across places and within places over time. This paper proposes a quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places. We show that a series of historical economic and demographic measures (proxies for variation in modes of settler colonial extraction and exploitation) are closely related to trajectories of state formation across places, measured as employment levels in core social policy system sectors.

@@

Historians tell us these projects were active locally. 1. Frontier; 2. Jim Crow; 3. Metropole. Can we use demographic, legal, and employment data to map it out / model it?

@@

Stratification in frontier was settler social democracy; flat social relations among settlers, extreme hostility to Native. Stratification in Jim Crow was rigid racial + class hierarchy; extreme wealth concentration and rigid racialized exploitation + social exclusion. Stratification in metropole was more fluid racial + class hierarchy; extreme concentration but porous boundaries, fewer bright lines.

Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development. This can lock in particular racial regimes over time as social policy does work on social groups.

@@

IMPLICATIONS - a) settler colonization as structure / structuring: context is key, and when SC is active, likely can be predictive; b) durable consequences for policy institutions (legacies, etc); c) comparative state formation / comparative welfare / carceral states.

# What is already known about settler colonization and state formation in late 19th and Early 20th c.

## Frontier

review here

## Jim Crow

review here

## Metropole

review here

@@

# methods

@@

Less interested in causal identification, more interested in understanding spatial patterns and trajectories. We already know there was spatial variation in settler colonial contexts and policy systems, trying to see if we can capture it systematically with national quant data. Move from specific to general.

demonstrate var(infrastructure), show e[infrastructure] = f(settler interests)

negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1

positive case: white immigrants -> more social workers in t1

here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.

## NOTES ON GEOGRAPHIES and MEASUREMENT

TALK ABOUT STATES, TERRITORIES, WESTERN EXPANSION. CONTESTATION AND TRANSFORMATION ARE KEY PROCESSES OF COLONIZATION. SETTLER COLONIALISM IS A SPATIAL STRUCTURE, NOT A SETTLED FACT, ESP VISIBLE IN THIS PERIOD.

STATES ARE WEIRD UNITS, BUT MAP CLEANLY ONTO ADMIN / LEGAL UNITS. TRIBAL NATIONS BOUNDARIES ARE ALSO V USEFUL, AS ARE SUB-STATE ANALYSES (EG CITIES COUNTIES ETC).

geography itself endogenous to settler colonization, this is tricky!

## sources of data

Census; ARCIA; Interior; etc

CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho...

we treat police, education, and social work employment as core areas of the domestic social state. what is the name for that? the welfare state? the carceral state? the therapeutic state? that thing.

let's look at how it played out in the progressive era across states and regions. we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression.

First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita?

@@

Figure \@ref(fig:blk-map) shows the distribution of the recorded Black population across US states in 1920. Panel 1 displays this data as the relative population composition of Black residents relative to to all residents, and panel 2 displays the absolute size of the Black population across US states.

DISCUSS THE DISTRIBUTION

@@

next let's look at sharecropping

@@

Places with large Black pops in 1920 also had very high sharecropping rates. Not a perfect correlation, VA and FL are negative off-trend, and lots of variance for low Black pop states, probably just measurement error though. Likely makese sense to use sharecropping as log count rather than as proportion.

>>>>

# Introduction

<!-- HOW ARE PROJECTS / PROBLEMS OF SETTLER COLONIAL GOVERNANCE RESOLVED? WHAT ORGS AND INSTITUTIONS ARE BUILT?  -->

Settler colonial regimes seek to resolve different problems across places and within places over time. This paper proposes a quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places (simultaneous settler projects). We show that a series of historical economic and demographic measures (proxies for variation in modes of settler colonial extraction and exploitation) are closely related to trajectories of state formation across places, measured as employment levels in core social policy system sectors.

@@

<!-- Historians tell us these projects were active locally. 1. Frontier; 2. Jim Crow; 3. Metropole. Can we use demographic, legal, and employment data to map it out / model it? -->

@@

<!-- Stratification in frontier was settler social democracy; flat social relations among settlers, extreme hostility to Native. Stratification in Jim Crow was rigid racial + class hierarchy; extreme wealth concentration and rigid racialized exploitation + social exclusion. Stratification in metropole was more fluid racial + class hierarchy; extreme concentration but porous boundaries, fewer bright lines. -->

<!-- Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development. This can lock in particular racial regimes over time as social policy does work on social groups. -->

@@

<!-- IMPLICATIONS - a) settler colonization as structure / structuring: context is key, and when SC is active, likely can be predictive; b) durable consequences for policy institutions (legacies, etc); c) comparative state formation / comparative welfare / carceral states.  -->

<!-- # What is already known about settler colonization and state formation in late 19th and Early 20th c.  -->

<!-- ## Frontier -->

<!-- review here -->

<!-- ## Jim Crow -->

<!-- review here -->

<!-- ## Metropole  -->

<!-- review here -->

@@

# Methods

@@

<!-- Less interested in causal identification, more interested in understanding spatial patterns and trajectories. We already know there was spatial variation in settler colonial contexts and policy systems, trying to see if we can capture it systematically with national quant data. Move from specific to general.  -->

Generally our approach demonstrates suggests that state formation in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place.

\[E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})\]

<!-- negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1 -->

<!-- positive case: white immigrants -> more social workers in t1 -->

<!-- here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.  -->

<!-- ## NOTES ON GEOGRAPHIES and MEASUREMENT -->

<!-- TALK ABOUT STATES, TERRITORIES, WESTERN EXPANSION. CONTESTATION AND TRANSFORMATION ARE KEY PROCESSES OF COLONIZATION. SETTLER COLONIALISM IS A SPATIAL STRUCTURE, NOT A SETTLED FACT, ESP VISIBLE IN THIS PERIOD.  -->

<!-- STATES ARE WEIRD UNITS, BUT MAP CLEANLY ONTO ADMIN / LEGAL UNITS. TRIBAL NATIONS BOUNDARIES ARE ALSO V USEFUL, AS ARE SUB-STATE ANALYSES (EG CITIES COUNTIES ETC).  -->

<!-- geography itself endogenous to settler colonization, this is tricky! -->

## Sources of data

Census (pop, employment, group quarters); ARCIA; Interior; BLM; US Forest Service.

<!-- CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho... -->

<!-- we treat police, education, and social work employment as core areas of the domestic social state. what is the name for that? the welfare state? the carceral state? the therapeutic state? that thing.  -->

<!-- let's look at how it played out in the progressive era across states and regions. we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression. -->

<!-- First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita? -->

@@

Figure \@ref(fig:blk-map) shows the distribution of the recorded Black population across US states in 1920. Panel 1 displays this data as the relative population composition of Black residents relative to to all residents, and panel 2 displays the absolute size of the Black population across US states. We display panel 2 on a logarithmic scale to enable closer examination of the first stage of the Great Migration. While Black populations were a low relative share of the population in places like New York, Pennsylvania and Illinois in 1920, there were substantial and growing Black populations in these states, as shown on the count, rather than proportion, scale.

@@

Next let's look at sharecropping

First, I evaluate the bivariate relationship between Black farm residency and Black tenant farm residency. The gap between the line y=x and plotted values indicates the number of Black people living on farms owned by their family. Tenant farming was the dominant mode of farming for Black farmers in all states with a substantial population of Black farmers in 1920. Virginia appears to be the only state with a substantial share of Black farm owners as of 1920. In states like Georgia, Mississippi, and South Carolina, these data clearly illustrate the dominance of the tenant-farming based sharecropping system.

@@

Places with large Black pops in 1920 also had very high sharecropping rates. Not a perfect correlation, VA and FL are negative off-trend, and lots of variance for low Black pop states, probably just measurement error though. Likely makes sense to use sharecropping as log count rather than as proportion.
